---
title: "Appendix"
subtitle: "Signal or noise? Minimising errors in image-based AI for landscape-scale marine ecosystem monitoring"
author:
  - name: Mathew Wyatt
    email: m.wyatt@aims.gov.au
    affiliations:
      - ref: aimswa
      - ref: owa
  - name: Julie Vercelloni
    corresponding: true
    affiliations:
      - ref: aimsqld
      - ref: qut
  - name: Cal Faubel
    affiliations:
     - ref: aimswa
  - name: Jamie Colquhuon
    affiliations:
     - ref: aimswa
  - name: Mary Wakeford
    affiliations:
     - ref: aimswa
  - name: Shaun Wilson
    affiliations:
     - ref: aimswa
  - name: Christopher J. Fulton
    affiliations:
      - ref: aimswa
      - ref: owa
affiliations:
  - id: owa
    name: Oceans Institute, University of Western Australia, Crawley, Western Australia 6009
  - id: aimswa
    name: Australian Institute of Marine Science, Crawley, Western Australia 6009
  - id: aimsqld
    name: Australian Institute of Marine Science, Cape Cleveland, Queensland 4810
  - id: qut
    name: Centre for Data Science, Queensland University of Technology, Brisbane, Queensland 4000
engine: knitr
format:
  html:
    toc: true
    toc-location: left
    title-block-banner: true
    toc-depth: 3
    highlight-style: atom-one
    embed-resources: true
    theme:
      light: flatly
      dark: darkly
    code-overflow: wrap
    code-fold: false
    number-sections: true
    number-depth: 2
    shift-heading-level-by: -1
    crossref:
      lst-title: Code listing
    fig-align: center
    text-align: center
    acm-html: default
execute: 
  message: false
  warning: false
  cache: true
editor:
  markdown:
    wrap: 72
---

```{r setup, chunkOpts, echo = FALSE}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE)
#knitr::opts_knit$set(root.dir = "..")
```

### Persistent bias and noise

```{r, fig.width=10, fig.height=8}
#| include: true
#| echo: false
#| eval: true
#| label: fig-sd
#| fig-cap: Estimates of main benthic habitat abundance within each of eight *Sargassum* meadow sites with annotator calibration. Obtained with image-based analyses with averaged cover values across individual observers. The group “Other” is introduced to constrain the sum to 100% cover for visualization purposes. Error bars show standard deviations across observers with the number of observers per class indicated in the coloured boxes.  
knitr::include_graphics(file.path(here::here(), "appendix/figures", "crowd_mean_sd.png"), rel_path = FALSE)
```

```{r, fig.width=10, fig.height=10}
#| include: true
#| echo: false
#| eval: true
#| label: fig-chord
#| fig-cap: Chord diagrams for observers 1–5 (a–e). For each observer, links represent how their “Can’t Tell” responses correspond to classifications made by the other observers for the same patches. Numbers indicate the total number of patches assessed.    
knitr::include_graphics(file.path(here::here(), "appendix/figures", "chord_diagrams_all.png"), rel_path = FALSE)
```

### Statistical modelling 
 
Bias and pattern noise are explored using multivariate analysis, specifically through the application of Generalized Linear Latent Variable Models (GLLVMs). 

Two independent GLLVMs were fitted using a beta distribution to analyse benthic community composition and attribute differences to collection method and *in situ* field experience, respectively. The models present the same structure with method and field experience treated as a categorical predictor, with line transects and file experience set as the reference levels. The models include a random effects structure at the levels of sites and observers. Significance was determined by identifying benthic communities whose confidence intervals did not include zero.

Formally, a beta distribution is used to model the community composition for community $j$ at site $i$. 

$$
\begin{align*}
Y_{ij} &\sim \text{Beta}(\mu_{ij}, \phi_j) \\
\text{logit}(\mu_{ij}) &= \beta_{0j} + \beta_{1j} \cdot \text{X}_i + u^{\text{site}}_{ij} + u^{\text{obs}}_{ij} + \mathbf{z}_i^\top \boldsymbol{\lambda}_j \\
u^{\text{site}}_{ij} &\sim \mathcal{N}(0, \sigma^2_{\text{site},j}) \\
u^{\text{obs}}_{ij} &\sim \mathcal{N}(0, \sigma^2_{\text{obs},j}) \\
\mathbf{z}_i &\sim \mathcal{N}(\mathbf{0}, \mathbf{I}_2)
\end{align*}
$$

The model estimates the logit-transformed mean cover $\mu_{ij}$ of community $j$ at site $i$, where $\beta_{0j}$ is the intercept for class $j$, $\beta_1j$ captures the fixed effect of method and field experiences,  $u^{\text{site}}_{ij}$ represents site-specific random intercepts, and $u^{\text{obs}}_{ij}$ is a random intercept capturing variation across observers. $\mathbf{z}_i$ is a vector of two latent variables for observation $i$, shared across all classes and $\boldsymbol{\lambda}_j$ is a vector of loadings (one per latent variable) specific to class $j$. The precision parameter $\phi_j$ for the beta distribution was fixed at 1 for all communities or groups. Reproducible R code for the GLLVMs is available in `scripts/gllvm.R`. 

#### Model diagnostics 

For both models, the goodness-of-fit was poorer for less abundant classes. Additionally, the null model (excluding the fixed effect) had slightly lower AIC values compared to the model that included the covariate.

##### Bias associated with method collection 

```{r, fig.width=10, fig.height=8}
#| include: true
#| echo: false
#| eval: true
#| label: fig-resid1
#| fig-cap: Model residuals.    
knitr::include_graphics(file.path(here::here(), "appendix/figures", "residm1.png"), rel_path = FALSE)
```

```{r, fig.width=10, fig.height=8}
#| include: true
#| echo: false
#| eval: true
#| label: fig-fit1
#| fig-cap: Model goodness-of-fit.    
knitr::include_graphics(file.path(here::here(), "appendix/figures", "fitm1.png"), rel_path = FALSE)
```

##### Pattern noise

```{r, fig.width=10, fig.height=8}
#| include: true
#| echo: false
#| eval: true
#| label: fig-resid2
#| fig-cap: Model residuals.    
knitr::include_graphics(file.path(here::here(), "appendix/figures", "residm2.png"), rel_path = FALSE)
```

```{r, fig.width=10, fig.height=8}
#| include: true
#| echo: false
#| eval: true
#| label: fig-fit2
#| fig-cap: Model goodness-of-fit.    
knitr::include_graphics(file.path(here::here(), "appendix/figures", "fitm2.png"), rel_path = FALSE)
```

